Estimating Uncertainty for Tipping Points with Deep Learning

### 2022-05-11 manuscript/manuscript.Rmd
plots

```{r}
hopf |> 
  group_by(t,type,variable) |> 
  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
  rename(species = variable) |> 
  mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
  ggplot(aes(t, col=type)) + 
    geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
    geom_line(aes(y=mean)) +
    #geom_vline(aes(xintercept = 100)) + 
    facet_wrap(~species, ncol=1) +
    theme_bw() + 
    ggtitle("A. Hopf bifurcation")
```

```{r}
panelB <- sn |>
  group_by(t,type,variable) |> 
  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
  rename(species = variable) |>
  ggplot(aes(t, col=type)) + 
    geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
    geom_line(aes(y=mean)) +
    #geom_vline(aes(xintercept = 100)) + 
    facet_wrap(~species, ncol=1) +
    theme_bw()

panelC <- stoch |>
  group_by(t,type,variable) |> 
  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
  rename(species = variable) |>
  ggplot(aes(t, col=type)) + 
    geom_ribbon(aes(ymin = pmax(mean-2*sd,0), ymax = mean+2*sd, fill=type), alpha=0.5) +
    geom_line(aes(y=mean)) +
    #geom_vline(aes(xintercept = 100)) + 
    facet_wrap(~species, ncol=1) +
    theme_bw() + 
  ggtitle("C. Stochastic transition")

panelA / panelB / panelC
```

```{r}
mcmc_scores <- bind_rows(
  read_csv("../greta/scores/scores_saddlenode.csv.gz") |> mutate(scenario="saddlenode"),
read_csv("../greta/scores/scores_stochastic.csv.gz"),
read_csv("../greta/scores/scores_hopf.csv.gz") |> 
  filter(scenario == "hopf_increasing") |>
  mutate(scenario = forcats::fct_recode(scenario, hopf = "hopf_increasing"),
         reps = 1)
) |> rename(method=model)
mcmc_scores |> 
  pivot_longer(c(logs, crps), names_to = "metric", values_to = "score") |> 
  ggplot(aes(method, score, col=variable)) + geom_boxplot() + 
  facet_grid(metric~scenario, scales="free_y")
```

>>>>

```{r}

forecasts <- read_parquet("data/forecasts.parquet")
observations <- read_parquet("data/observations.parquet")

hopf_obs <- observations |> filter(simulation=="hopf")
stochastic_obs <- observations |> filter(simulation=="stochastic")
saddle_obs <- observations |> filter(simulation=="saddle")
```

```{r fig.cap="Forecasts of the Hopf bifurcation under each model"}
forecasts |> filter(simulation=="hopf", reps == 1, group == 1) |> 
  ggplot(aes(t, col=type, fill=type)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5, data = hopf_sims) +
  geom_line(aes(y=mean)) +
  geom_line(aes(y=mean), data = hopf_obs) +
  facet_grid(forecasting_model ~ variable, scales = "free") +
  theme_bw() + 
  ggtitle("Hopf bifurcation")
```

```{r}
forecasts |> filter(simulation=="saddle", group == 1, reps == 1) |> 
  ggplot(aes(t, col=type, fill=type)) + 
  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = saddle_sims) +
  geom_line(aes(y=mean)) +
  geom_line(aes(y=mean), data = saddle_obs) +
  facet_wrap(~forecasting_model, scales = "free") +
  #facet_grid(forecasting_model ~ reps, scales = "free") +
  theme_bw() + 
  ggtitle("Saddle-Node bifurcation")
```

```{r}
forecasts |> filter(simulation=="stochastic", group == 1, reps==1) |> 
  ggplot(aes(t, col=type, fill=type)) + 
  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = stochastic_sims) +
  geom_line(aes(y=mean)) +
  geom_line(aes(y=mean), data = stochastic_obs) +
  facet_wrap(~forecasting_model, scales = "free") +
  #facet_grid(forecasting_model ~ reps, scales = "free") +
  theme_bw() + 
  ggtitle("Stochastic transition")
```

```{r}
scores <- arrow::open_dataset("../scores", format="csv") |> collect()

# shift logs score to strictly positive, non-infinite values so we can use a log-scale
shift <- scores |> 
  filter(!is.infinite(logs)) |> 
  mutate(logs = logs - min(logs) +.01) |>
  pivot_longer(c(logs, crps), 
               names_to = "metric", 
               values_to = "score")
```

```{r}
shift |> 
  ggplot(aes(forecasting_model, score, color=forecasting_model, fill=forecasting_model)) + 
  geom_violin() +
  facet_grid(metric ~ simulation, scales = "free") + 
  scale_y_log10() +
  theme_bw() + 
  coord_flip()

  ggplot(aes(simulation, score, color=forecasting_model, fill=forecasting_model)) + 
  geom_violin() +
  facet_wrap(~ metric, scales = "free", ncol=1) + 
  scale_y_log10() + 
  theme_bw()
```

```{r}
shift |>
  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
  geom_jitter(alpha = 0.01, show.legend = FALSE) +
  facet_grid(metric ~ simulation, scales = "free") + 
  scale_y_log10() +
  theme_bw() + 
  coord_flip()

```{r}
# scores over time, averaged over replicates
over_time <- shift |> 
  group_by(t, simulation,forecasting_model, metric ) |> 
  summarise(score = mean(score))

over_time |> 
  ggplot(aes(t, score, col = forecasting_model)) + 
  geom_point(size=1) +
  facet_grid(metric ~ simulation,
             scales="free") + 
  theme_bw() + scale_y_log10()
